---
title: "Predicting Alcoholism Using Various Characteristics — Analysis Code"
subtitle: "STATS 101C Final Project (cleaned & annotated)"
author: "Nikita Park, Jeremy Shiu, Ian Turner, Mingchen Wang, Josh Xu"
date: "2023-12-17"
output: pdf_document
---

```{r setup, include=FALSE}
# Show code and results, but keep messages/warnings out of the knitted report.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# A single global seed keeps every random step (imputation draws, resampling,
# train/test splits) reproducible across runs.
GLOBAL_SEED <- 123
set.seed(GLOBAL_SEED)
```

> **What this file does.** It walks through the full modeling pipeline described
> in the write-up: exploratory analysis, handling the large amount of missing
> data via multiple imputation, a sequence of increasingly capable models, and
> the final tuned XGBoost model that scored 73.38% on the Kaggle leaderboard.
>
> **Reproducibility note.** Several steps (multiple imputation, bootstrap tuning)
> are lengthy to run. Those steps are written out in full but wrapped so the code
> loads a previously saved result when one exists, and only recomputes when it
> does not. Look for the `if (file.exists(...))` guards.

# 0. Setup

```{r libraries}
# --- Data handling & plotting ---
library(ggplot2)   # EDA density / bar plots
library(dplyr)     # light data wrangling

# --- Missing-data assessment & imputation ---
library(VIM)       # visualize the pattern of missingness (Figure 2)
library(Hmisc)     # aregImpute(): the imputation method we ultimately chose
library(Amelia)    # bootstrap-EM imputation, used for the test set
library(missForest)# random-forest imputation (one of the methods we compared)

# --- Models ---
library(MASS)      # stepAIC() for stepwise logistic regression
library(class)     # knn()
library(xgboost)   # gradient-boosted trees: our final model
library(caret)     # train(): parameter tuning via resampling
library(DiagrammeR)# render individual XGBoost trees
```

```{r load-data}
# Kaggle-provided data.
#   train: 70,000 patients x 27 columns (26 predictors + Alcoholic.Status)
#   test : 30,000 patients x 26 columns (Alcoholic.Status withheld for scoring)
# Predictors mix vital signs (age, sex, blood pressure, cholesterol, liver
# enzymes, ...) with lifestyle factors (smoking status).
train_raw <- read.csv("TrainSAData2.csv")
test_raw  <- read.csv("TestSAData2NoY.csv")

# Column groupings reused throughout. Everything except the ID, the response,
# and the six categorical fields is treated as numeric.
response_col    <- "Alcoholic.Status"
id_col          <- "ID"
categorical_cols <- c("sex", "hear_left", "hear_right",
                      "BMI.Category", "AGE.Category", "Smoking.Status")
```

# 1. Exploratory Data Analysis

```{r class-balance}
# The response is almost perfectly balanced, so plain accuracy is a fair metric
# and no class weighting / resampling for imbalance is required.
#   Reported in write-up: 34,887 alcoholics (49.84%) vs 35,113 non (50.16%).
class_balance <- prop.table(table(train_raw$Alcoholic.Status))
class_balance
```

```{r eda-numeric-density, fig.height=6}
# Density of each strong numeric predictor, split by alcoholic status, to see
# whether the two groups separate. Hemoglobin and gamma_GTP show the clearest
# shift; height and weight separate mainly because they track sex.
strong_numeric <- c("height", "weight", "hemoglobin", "gamma_GTP")

for (v in strong_numeric) {
  print(
    ggplot(train_raw, aes(x = .data[[v]], color = Alcoholic.Status)) +
      geom_density() +
      labs(title = paste("Density of", v, "by alcoholic status"), x = v)
  )
}
```

```{r eda-categorical-bars, fig.height=4}
# Stacked bar plots for the categorical predictors. Sex and smoking status are
# the most visually predictive: men and current smokers skew alcoholic.
strong_categorical <- c("sex", "AGE.Category", "Smoking.Status")

for (v in strong_categorical) {
  print(
    ggplot(train_raw, aes(x = .data[[v]], fill = Alcoholic.Status)) +
      geom_bar() +
      labs(title = paste("Alcoholic status by", v), x = v)
  )
}
```

```{r eda-single-predictor-baseline}
# Sanity check: how far does the single strongest predictor get us on its own?
# A logistic regression on hemoglobin alone scored ~63.93% on the test set,
# confirming that no single variable is sufficient and a multivariable model is
# needed. (Fit on complete cases since hemoglobin has missing values.)
hemoglobin_baseline <- glm(
  factor(Alcoholic.Status) ~ hemoglobin,
  data   = train_raw,
  family = "binomial"
)
summary(hemoglobin_baseline)$coefficients
```

# 2. Missing Data and Multiple Imputation

The dataset's biggest obstacle was a large, roughly-random scattering of missing
values across nearly every predictor. Because the missingness was spread out, we
could not drop rows or columns without losing most of the data, so we imputed.
We compared four packages — **Amelia**, **Hmisc (`aregImpute`)**, **mice**, and
**missForest** — and the models trained on `aregImpute`'s output scored highest.

```{r missingness-plot, fig.height=6}
# Figure 2: how many values are missing per predictor. Every predictor except
# ID and the response is missing on the order of ~5,000 values; AGE.Category is
# the worst. This picture is what ruled out list-wise deletion.
missing_summary <- aggr(
  train_raw,
  numbers = TRUE, sortVars = TRUE, cex.axis = 0.6, gap = 3,
  ylab = c("Missing data", "Pattern")
)
```

```{r aregimpute-train}
# aregImpute() = additive-regression + bootstrap + predictive-mean-matching
# multiple imputation. We use predictive-mean matching ("pmm") because the
# alternative (regression) can't handle the categorical predictors, and
# n.impute = 5 as a compromise between imputation quality and runtime.
#
# This fit takes a very long time, so it is cached to disk. Delete
# "aregimpute_kaggle_train.RData" to force a fresh run.
areg_cache <- "aregimpute_kaggle_train.RData"

if (file.exists(areg_cache)) {
  areg_fit <- readRDS(areg_cache)
} else {
  set.seed(GLOBAL_SEED)
  areg_fit <- aregImpute(
    Alcoholic.Status ~ sex + age + height + weight + waistline +
      sight_left + sight_right + hear_left + hear_right + SBP + DBP + BLDS +
      tot_chole + HDL_chole + LDL_chole + triglyceride + hemoglobin +
      urine_protein + serum_creatinine + SGOT_AST + SGOT_ALT + gamma_GTP +
      BMI + BMI.Category + AGE.Category + Smoking.Status,
    data = train_raw, n.impute = 5
  )
  saveRDS(areg_fit, file = areg_cache)
}
```

```{r build-imputed-training-set}
# Reconstruct a complete training frame by dropping aregImpute's imputed draws
# into the cells that were originally missing.
#
# We use the 2nd of the 5 multiple imputations throughout (any single draw is
# valid; we simply fix one for reproducibility).
IMPUTE_DRAW <- 2

train_imputed <- train_raw

# aregImpute stores categorical variables by their integer codes, so encode the
# observed categoricals to match before filling in the imputed codes.
train_imputed$sex        <- ifelse(train_imputed$sex        == "Male",     2, 1)
train_imputed$hear_left  <- ifelse(train_imputed$hear_left  == "Abnormal", 1, 2)
train_imputed$hear_right <- ifelse(train_imputed$hear_right == "Abnormal", 1, 2)
train_imputed$BMI.Category   <- as.numeric(factor(train_raw$BMI.Category))
train_imputed$AGE.Category   <- as.numeric(factor(train_raw$AGE.Category))
train_imputed$Smoking.Status <- as.numeric(factor(train_raw$Smoking.Status))

# Fill every imputed variable in one pass instead of 25 near-identical lines:
# for each variable, overwrite exactly the rows aregImpute flagged as missing.
for (v in names(areg_fit$imputed)) {
  missing_rows <- areg_fit$na[[v]]
  train_imputed[missing_rows, v] <- areg_fit$imputed[[v]][, IMPUTE_DRAW]
}
```

```{r encode-imputed-training-set}
# Two export formats are used downstream:
#   * a purely numeric frame (tree models like the integer codes), and
#   * a human-readable frame with the category labels restored (used by the
#     logistic-regression models, which need real factors).
#
# The factor level orders below match R's default alphabetical ordering that
# as.numeric(factor(...)) assigned above, so the codes map back correctly.
train_labeled <- train_imputed
train_labeled$sex        <- c("Female", "Male")[train_labeled$sex]
train_labeled$hear_left  <- c("Abnormal", "Normal")[train_labeled$hear_left]
train_labeled$hear_right <- c("Abnormal", "Normal")[train_labeled$hear_right]
train_labeled$BMI.Category   <- c("Healthy", "Obese", "Overweight",
                                  "Underweight")[train_labeled$BMI.Category]
train_labeled$AGE.Category   <- c("Mid-aged", "Old", "Very Old",
                                  "Young")[train_labeled$AGE.Category]
train_labeled$Smoking.Status <- c("Never Smoked", "Still Smoking",
                                  "Used to Smoke")[train_labeled$Smoking.Status]

# Persist both versions so later chunks (and other scripts) can reload them
# without repeating the imputation.
write.csv(train_imputed,  "aregImpute_Train.csv",           row.names = FALSE)
write.csv(train_labeled,  "aregImpute_Train_character.csv", row.names = FALSE)
```

```{r amelia-impute-test}
# The test set needs imputing too. We use Amelia here (fast, bootstrap-EM) and
# tell it which columns are nominal so it doesn't treat categories as numbers.
# Result is cached for the same runtime reasons as the training imputation.
amelia_cache <- "amelia_test_data.csv"

if (file.exists(amelia_cache)) {
  test_imputed <- read.csv(amelia_cache)
} else {
  amelia_input <- test_raw[, -1]  # drop ID; it carries no information to impute
  for (v in categorical_cols) amelia_input[[v]] <- as.factor(amelia_input[[v]])

  set.seed(GLOBAL_SEED)
  amelia_out <- amelia(
    amelia_input, m = 1,
    noms = which(names(amelia_input) %in% categorical_cols)
  )

  # Re-attach the original IDs so predictions line up with the submission file.
  test_imputed <- cbind(ID = test_raw$ID, amelia_out$imputations$imp1)
  write.csv(test_imputed, amelia_cache, row.names = FALSE)
}
```

```{r submission-helper}
# Every model below produces a Kaggle submission in the same shape, so define
# the formatting once: probabilities -> "Y"/"N" at a 0.5 cutoff, paired with ID.
make_submission <- function(probabilities, ids, file) {
  prediction <- ifelse(probabilities >= 0.5, "Y", "N")
  submission <- data.frame(ID = ids, Alcoholic.Status = prediction)
  write.csv(submission, file = file, row.names = FALSE)
  invisible(submission)
}
```

# 3. Baseline Models: Logistic Regression

Before reaching for tree ensembles we established logistic-regression baselines,
including stepwise variable selection by both BIC and AIC. Linear and quadratic
discriminant analysis (LDA / QDA) were also tried but scored poorly even on the
training data — consistent with the classes not being linearly separable — so
they were dropped from contention and are not reproduced here.

```{r glm-full-and-stepwise}
# Full logistic-regression model on the labeled, fully-imputed training data.
glm_data <- read.csv("aregImpute_Train_character.csv")

# Encode the response as 0/1 and make sure categoricals are treated as factors.
glm_data$Alcoholic.Status <- ifelse(glm_data$Alcoholic.Status == "N", 0, 1)
for (v in categorical_cols) glm_data[[v]] <- as.factor(glm_data[[v]])

glm_full <- glm(
  Alcoholic.Status ~ sex + age + height + weight + waistline + sight_left +
    sight_right + hear_left + hear_right + SBP + DBP + BLDS + tot_chole +
    HDL_chole + LDL_chole + triglyceride + hemoglobin + urine_protein +
    serum_creatinine + SGOT_AST + SGOT_ALT + gamma_GTP + BMI + BMI.Category +
    AGE.Category + Smoking.Status,
  data = glm_data, family = "binomial"
)

# Stepwise selection. BIC (k = log n) penalizes complexity more heavily than
# AIC, so it yields a smaller model. Both are cached because step() is slow.
if (file.exists("aregimpute_back_BIC.RData")) {
  glm_backward_bic <- readRDS("aregimpute_back_BIC.RData")
} else {
  glm_backward_bic <- step(glm_full, direction = "backward",
                           k = log(nrow(glm_data)))
  saveRDS(glm_backward_bic, "aregimpute_back_BIC.RData")
}

if (file.exists("aregimpute_back_AIC.RData")) {
  glm_backward_aic <- readRDS("aregimpute_back_AIC.RData")
} else {
  glm_backward_aic <- stepAIC(glm_full, direction = "backward")
  saveRDS(glm_backward_aic, "aregimpute_back_AIC.RData")
}
```

```{r glm-predict}
# Score both stepwise models on the imputed test set. type = "response" is
# essential: it returns probabilities in [0, 1] rather than raw log-odds, so the
# 0.5 cutoff in make_submission() is meaningful.
bic_prob <- predict(glm_backward_bic, newdata = test_imputed, type = "response")
aic_prob <- predict(glm_backward_aic, newdata = test_imputed, type = "response")

make_submission(bic_prob, test_imputed$ID, "prototype2.csv")  # backward-BIC GLM
make_submission(aic_prob, test_imputed$ID, "prototype3.csv")  # backward-AIC GLM
```

# 4. Final Model: XGBoost

XGBoost builds trees sequentially, each one fitting the residual errors of the
ensemble so far (gradient boosting). It was our best algorithm, so we spent the
bulk of the project tuning it.

```{r xgboost-matrices}
# XGBoost needs a purely numeric feature matrix and a 0/1 label. Work from the
# numeric imputed training set and drop the ID column.
xgb_train <- read.csv("aregImpute_Train.csv")
xgb_train <- xgb_train[, setdiff(names(xgb_train), id_col)]

xgb_label   <- as.integer(xgb_train$Alcoholic.Status == "Y")
xgb_feature <- data.matrix(xgb_train[, setdiff(names(xgb_train), response_col)])
```

```{r xgboost-initial}
# Default XGBoost: no tuning, no pruning. Establishes the tree-model baseline.
#   Reported test accuracy: ~72.28% (already better than every earlier model).
set.seed(GLOBAL_SEED)
xgb_initial <- xgboost(
  data      = xgb_feature,
  label     = xgb_label,
  nrounds   = 100,
  objective = "binary:logistic",
  eval_metric = "logloss",
  verbose   = 0
)
```

```{r xgboost-importance, fig.height=6}
# Variable importance (Figure 5). The standout result: sex is by far the most
# important predictor, followed by age, gamma_GTP, and smoking status.
xgb_importance <- xgb.importance(model = xgb_initial)
print(xgb_importance)
xgb.plot.tree(model = xgb_initial, trees = 0)  # inspect the first tree
```

```{r xgboost-tuning}
# Tuning via caret::train(). The write-up found that:
#   * resampling (10-fold CV ~73.23%, then bootstrap ~73.33%) beat a single fit,
#   * pruning max_depth to 2 helped, and
#   * the default learning rate eta = 0.3 was hard to beat.
# The grid below is centered on those findings. This is the most expensive step
# in the file, so it is cached; delete the .RData to re-tune.
xgb_tune_cache <- "xgboost_tuned_caret.RData"

if (file.exists(xgb_tune_cache)) {
  xgb_tuned <- readRDS(xgb_tune_cache)
} else {
  tune_grid <- expand.grid(
    max_depth        = c(1, 2, 3, 4),  # 2 was best (see Figure 4)
    eta              = c(0.3, 0.4),    # 0.3 (default) won
    nrounds          = c(50, 100, 150, 200),
    gamma            = 0,
    colsample_bytree = 0.8,
    min_child_weight = 1,
    subsample        = c(0.5, 0.6667, 0.8333, 1.0)
  )

  set.seed(GLOBAL_SEED)
  xgb_tuned <- train(
    x = xgb_feature,
    y = factor(ifelse(xgb_label == 1, "Y", "N")),
    method    = "xgbTree",
    metric    = "Accuracy",
    trControl = trainControl(method = "boot", number = 25),  # bootstrap resampling
    tuneGrid  = tune_grid
  )
  saveRDS(xgb_tuned, xgb_tune_cache)
}

# Best parameter combination found by the search.
xgb_tuned$bestTune
```

```{r xgboost-variable-selection}
# We also tested dropping the weakest 20, 10, 5, 2, and 1 predictor(s) by
# importance to combat overfitting. Every reduced model scored *worse* on the
# leaderboard, so the final model keeps all predictors. (This is a
# leaderboard-chasing choice; a real study would drop redundant predictors such
# as BMI.Category and the eyesight/hearing fields — see the Discussion.)
#
# Illustrative sweep, left un-run by default because the tuned model above is
# the deliverable. Set eval = TRUE to reproduce the experiment.
```

```{r xgboost-variable-selection-demo, eval=FALSE}
ranked_features <- xgb_importance$Feature  # most -> least important
for (n_drop in c(20, 10, 5, 2, 1)) {
  keep <- setdiff(ranked_features, tail(ranked_features, n_drop))
  set.seed(GLOBAL_SEED)
  reduced <- xgboost(
    data = xgb_feature[, keep], label = xgb_label,
    nrounds = 100, objective = "binary:logistic",
    eval_metric = "logloss", verbose = 0
  )
  cat("Dropped", n_drop, "weakest predictor(s):",
      1 - min(reduced$evaluation_log$train_logloss), "\n")
}
```

```{r xgboost-final-submission}
# Final predictions: align the test features to the training feature order,
# score with the tuned model, and write the submission that scored 73.38%.
xgb_test <- test_imputed
for (v in categorical_cols) xgb_test[[v]] <- as.integer(as.factor(xgb_test[[v]]))
xgb_test_matrix <- data.matrix(xgb_test[, colnames(xgb_feature)])

# caret's train object predicts class labels directly for xgbTree.
xgb_final_class <- predict(xgb_tuned, newdata = xgb_test_matrix)
xgb_final_prob  <- ifelse(xgb_final_class == "Y", 1, 0)
make_submission(xgb_final_prob, test_imputed$ID, "prototype4.csv")
```

# 5. Other Algorithms Explored

For completeness we also evaluated CatBoost, LightGBM, random forest, and
logistic regression with PCA-reduced predictors. None beat XGBoost (see the
model-comparison figure in the write-up). A k-nearest-neighbors model and an
artificial neural network were also attempted; KNN underperformed and the ANN
was too computationally expensive to finish. The KNN experiment is reproduced
below as a representative example; the others followed the same fit-then-compare
pattern and are described in the report rather than rerun here.

```{r knn-experiment}
# KNN on the numeric imputed data. Distance-based methods need the ID, response,
# and redundant category columns removed, and benefit from an internal
# validation split to gauge accuracy before touching the test set.
knn_train_full <- read.csv("aregImpute_Train_character.csv")
knn_label      <- as.numeric(factor(knn_train_full$Alcoholic.Status))

# Drop ID (1), response (2) and the categorical columns so only comparable
# numeric features remain.
drop_idx  <- c(1, 2, 9, 10, 25, 26, 27, 28)
knn_features <- knn_train_full[, -drop_idx]

# 70/30 internal split to estimate accuracy.
set.seed(113355)
train_idx <- sample(seq_len(nrow(knn_features)),
                    size = 0.7 * nrow(knn_features), replace = FALSE)

knn_pred <- knn(
  train = knn_features[train_idx, ],
  test  = knn_features[-train_idx, ],
  cl    = knn_label[train_idx],
  k     = 10
)

# Validation accuracy — low relative to XGBoost, which is why KNN was dropped.
knn_accuracy <- mean(knn_pred == knn_label[-train_idx])
knn_accuracy
```

# 6. Summary

The pipeline above reflects the reported results: logistic-regression baselines
in the ~64–72% range, a default XGBoost at ~72.28%, and a bootstrap-tuned
XGBoost pruned to `max_depth = 2` reaching 73.38% — good for 5th place overall
and 4th within the lecture. The key practical lessons were that intelligent
imputation (`aregImpute`) mattered more than model choice at the margin, and
that a smaller two-tree model both generalized better and ran faster than deeper
alternatives.
