---
title: "stat c173 project"
author: "Jeremy Shiu"
date: "2025-03-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Original data

The committed `final_data.csv` (one row per monitoring station) was produced
from the full raw Athens hourly dataset (`athens_data.csv`, ~439 MB, not stored
in this repo — see README). To let this document knit from the repo alone, the
chunk below reads the small `data/sample/athens_data.sample.csv` and writes a
demo file rather than overwriting the committed `final_data.csv`. To reproduce
the real `final_data.csv`, point `read_csv()` at the full `athens_data.csv`.

```{r}
library(dplyr)
library(readr)
data <- read_csv("data/sample/athens_data.sample.csv")

df_unique <- data %>% distinct(Latitude, Longitude, .keep_all = TRUE)
df_unique <- df_unique[,-c(1,4,18,19)]
df_unique <- df_unique[,-c(5,6,7,8,9,13,14)]

# write.csv(df_unique, "final_data.csv", row.names = FALSE)  # disabled: keeps committed final_data.csv intact
```


Final Data
```{r}
library(readr)
data <- read_csv("final_data.csv")
```

Non-spatial data analysis
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
```

Summary statistics
```{r}
summary(data$Latitude)
summary(data$Longitude)
summary(data$`Wind-Speed (U)`)
summary(data$`Wind-Speed (V)`)
summary(data$Temp)
summary(data$`Relative Humidity`)
summary(data$PM10)
```

Histograms
```{r}
hist(data$`Wind-Speed (U)`, main = "Longitudinal Wind Speed")
hist(data$`Wind-Speed (V)`, main = "Latitudinal Wind Speed")
hist(data$Temp, main = "Air Temperature")
hist(data$`Relative Humidity`, main = "Humidity")
hist(data$PM10, main = "PM10")
hist(data$O3, main = "Ozone Concentration")
```

ecdfs
```{r}
library(ggplot2)

ggplot(data, aes(x = `Wind-Speed (U)`)) +
  stat_ecdf(geom = "step", color = "blue") +
  labs(title = "Longitudinal Wind Speed ECDF", x = "Values", y = "ECDF") +
  theme_minimal()

ggplot(data, aes(x = `Wind-Speed (V)`)) +
  stat_ecdf(geom = "step", color = "blue") +
  labs(title = "Latitudinal Wind Speed ECDF", x = "Values", y = "ECDF") +
  theme_minimal()

ggplot(data, aes(x = `Temp`)) +
  stat_ecdf(geom = "step", color = "blue") +
  labs(title = "Air Temperature ECDF", x = "Values", y = "ECDF") +
  theme_minimal()

ggplot(data, aes(x = `Relative Humidity`)) +
  stat_ecdf(geom = "step", color = "blue") +
  labs(title = "Relative Humidity ECDF", x = "Values", y = "ECDF") +
  theme_minimal()

ggplot(data, aes(x = `PM10`)) +
  stat_ecdf(geom = "step", color = "blue") +
  labs(title = "PM10 ECDF", x = "Values", y = "ECDF") +
  theme_minimal()

ggplot(data, aes(x = `O3`)) +
  stat_ecdf(geom = "step", color = "blue") +
  labs(title = "Ozone Concentration ECDF", x = "Values", y = "ECDF") +
  theme_minimal()
```


```{r}
ggplot(data, aes(x = Longitude, y = Latitude)) +
  geom_point(color = "blue", size = 2) +
  labs(title = "Scatter Plot of Latitude vs Longitude",
       x = "Longitude",
       y = "Latitude") +
  theme_minimal()
```

Bubble Plot
```{r}
# Create a bubble plot for PM10 and locations
ggplot(data, aes(x = Longitude, y = Latitude, size = PM10)) +
  geom_point(alpha = 0.7, color = "blue") +  # Adjust transparency and color
  labs(title = "Bubble Plot of PM10 Concentrations by Location",
       x = "Longitude",
       y = "Latitude",
       size = "PM10 Concentration") +
  theme_minimal() +
  scale_size_continuous(range = c(3, 10))  # Adjust the range of bubble sizes
```

h-scatterplots
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(geosphere)

data <- read.csv("final_data.csv")

# Calculate pairwise distances between locations
dist_matrix <- distm(data[, c("Longitude", "Latitude")], fun = distHaversine)
dist_values <- as.vector(dist_matrix[upper.tri(dist_matrix)])
summary(dist_values)

# Define lag distance (e.g., 10 km)
lag_distance <- 10000  # in meters (10 km)

# Find pairs of points within the lag distance
pairs <- which(dist_matrix <= lag_distance & dist_matrix > 0, arr.ind = TRUE)

# Extract PM10 values for the pairs
pm10_pairs <- data.frame(
  PM10_1 = data$PM10[pairs[, 1]],
  PM10_2 = data$PM10[pairs[, 2]]
)

# Create the h-scatterplot
ggplot(pm10_pairs, aes(x = PM10_1, y = PM10_2)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = paste("H-Scatterplot for PM10 (Lag Distance =", lag_distance / 1000, "km)"),
       x = "PM10 at Location 1",
       y = "PM10 at Location 2") +
  theme_minimal()
```

```{r}
# Define lag distance (e.g., 10 km)
lag_distance <- 5000  # in meters (10 km)

# Find pairs of points within the lag distance
pairs <- which(dist_matrix <= lag_distance & dist_matrix > 0, arr.ind = TRUE)

# Extract PM10 values for the pairs
pm10_pairs <- data.frame(
  PM10_1 = data$PM10[pairs[, 1]],
  PM10_2 = data$PM10[pairs[, 2]]
)

# Create the h-scatterplot
ggplot(pm10_pairs, aes(x = PM10_1, y = PM10_2)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = paste("H-Scatterplot for PM10 (Lag Distance =", lag_distance / 1000, "km)"),
       x = "PM10 at Location 1",
       y = "PM10 at Location 2") +
  theme_minimal()
```

```{r}
# Define lag distance (e.g., 10 km)
lag_distance <- 20000  # in meters (10 km)

# Find pairs of points within the lag distance
pairs <- which(dist_matrix <= lag_distance & dist_matrix > 0, arr.ind = TRUE)

# Extract PM10 values for the pairs
pm10_pairs <- data.frame(
  PM10_1 = data$PM10[pairs[, 1]],
  PM10_2 = data$PM10[pairs[, 2]]
)

# Create the h-scatterplot
ggplot(pm10_pairs, aes(x = PM10_1, y = PM10_2)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = paste("H-Scatterplot for PM10 (Lag Distance =", lag_distance / 1000, "km)"),
       x = "PM10 at Location 1",
       y = "PM10 at Location 2") +
  theme_minimal()
```

Empirical Semivariogram
```{r}
# Load necessary libraries
library(gstat)
library(readr)
library(sp)
library(dplyr)

data <- read_csv("final_data.csv")
#data <- data %>%
#  filter(!is.na(PM10) & !is.na(Latitude) & !is.na(Longitude))

#data <- data %>%
#  filter(!is.na(O3))

# Convert to spatial data frame
coordinates(data) <- ~Longitude + Latitude
```

```{r}
# Compute Classical and Robust Empirical Semivariogram
vario_classical <- variogram(PM10 ~ 1, data)  # Classical
vario_robust <- variogram(PM10 ~ 1, data, cressie = TRUE)  # Robust Cressie-Hawkins estimator

# Inspect the empirical semivariogram to pick rough starting values
print(vario_classical)
max(vario_classical$gamma)  # use as rough psill target
max(vario_classical$dist)   # use ~1/3 of this as rough range target

# Fit Exponential Model
exp_model <- vgm(psill = 9000, model = "Exp", range = 15000, nugget = 1000)
exp_fit <- fit.variogram(vario_classical, exp_model, fit.method = 2)

# Fit Spherical Model
sph_model <- vgm(psill = 9000, model = "Sph", range = 20000, nugget = 1000)
sph_fit <- fit.variogram(vario_classical, sph_model, fit.method = 2)

# Fit Gaussian Model
gau_model <- vgm(psill = 9000, model = "Gau", range = 10000, nugget = 1000)
gau_fit <- fit.variogram(vario_classical, gau_model, fit.method = 2)

max_dist <- max(vario_classical$dist)

plot.new()
plot(vario_classical, 
     main = "Empirical Semivariogram with Fitted Models", 
     pch = 16, col = "black",
     ylim = c(0, 1.1*max(vario_classical$gamma)), xlim = c(0, max_dist)) # Adjust y-axis limits
lines(variogramLine(exp_fit, max_dist), col = "red", lwd = 2, lty = 2)
lines(variogramLine(sph_fit, (1/3) * max(vario_classical$dist)), col = "blue", lwd = 2, lty = 3)
lines(variogramLine(gau_fit, (1/3) * max(vario_classical$dist)), col = "green", lwd = 2, lty = 4)
legend("bottomright", legend = c("Exponential", "Spherical", "Gaussian"), 
       col = c("red", "blue", "green"), lwd = 2, lty = c(2, 3, 4))
```

```{r}
# Install and load necessary packages
#install.packages(c("gstat", "sp", "ggplot2", "dplyr"))
library(gstat)
library(sp)
library(ggplot2)
library(dplyr)

# Load the data
data <- read.csv("final_data.csv")

coordinates(data) <- ~Longitude + Latitude
sample_variogram <- variogram(PM10 ~ 1, data)
plot(sample_variogram)
variogram_ols <- fit.variogram(sample_variogram, vgm(model = "Sph"), fit.method=0)
plot(sample_variogram, model =variogram_ols, main = "OLS fit with spherical model")

## Weighted Least Squares (npairs)
variogram_npairs <- fit.variogram(sample_variogram, vgm(model="Sph"), fit.method = 1)
plot(sample_variogram, model = variogram_npairs, main = "WLS npairs fit with spherical model")

variogram_cressie <- fit.variogram(sample_variogram, vgm(model = "Sph"), fit.method=1)
plot(sample_variogram, variogram_cressie, main = "WLS Cressie fit with spherical model")
```

```{r}
exp_variogram_ols <- fit.variogram(sample_variogram, vgm(model = "Exp"), fit.method=0)
plot(sample_variogram, model =exp_variogram_ols, main = "OLS fit with exponential model")

## Weighted Least Squares (npairs)
exp_variogram_npairs <- fit.variogram(sample_variogram, vgm(model="Exp"), fit.method = 1)
plot(sample_variogram, model = exp_variogram_npairs, main = "WLS npairs fit with exponential model")

exp_variogram_cressie <- fit.variogram(sample_variogram, vgm(model = "Exp"), fit.method=1)
plot(sample_variogram, exp_variogram_cressie, main = "WLS Cressie fit with exponential model")
```


IDW interpolation
```{r}
# Load necessary libraries
library(gstat)
library(sp)

# Read the dataset
data <- read.csv("final_data.csv")

# Convert the data to a spatial object
#coordinates(data) <- ~Longitude + Latitude

# Define the bounding box of the data
x.range <- range(data$Longitude)
y.range <- range(data$Latitude)

# Create a grid for predictions
grd <- expand.grid(
  Longitude = seq(from = x.range[1], to = x.range[2], length.out = 100),
  Latitude = seq(from = y.range[1], to = y.range[2], length.out = 100)
)

# Convert the grid to a spatial object
coordinates(grd) <- ~Longitude + Latitude
gridded(grd) <- TRUE

# Perform IDW interpolation for PM10
idw.out <- idw(PM10 ~ 1, locations = ~Longitude + Latitude, data, grd)

# Collapse the vector of predicted values into a matrix
pred_matrix <- matrix(idw.out$var1.pred, 
                      nrow = length(seq(from = x.range[1], to = x.range[2], length.out = 100)),
                      ncol = length(seq(from = y.range[1], to = y.range[2], length.out = 100)))

# Raster map of the predictions
image(seq(from = x.range[1], to = x.range[2], length.out = 100),
      seq(from = y.range[1], to = y.range[2], length.out = 100),
      pred_matrix,
      xlab = "Longitude", ylab = "Latitude",
      main = "IDW Interpolation of PM10 Concentrations")

# Add observed data points
points(data, pch = 19, col = "black")

# Bubble plot of observed data
points(data, cex = data$PM10 / mean(data$PM10), pch = 19, col = "red")

# Add contours
contour(seq(from = x.range[1], to = x.range[2], length.out = 100),
        seq(from = y.range[1], to = y.range[2], length.out = 100),
        pred_matrix, add = TRUE, col = "black", labcex = 1)


```

```{r}
# Load necessary libraries
library(sp)
library(gstat)
library(ggplot2)

# Read the dataset
data <- read.csv("final_data.csv")

# Convert the data to a spatial object
coordinates(data) <- ~Longitude + Latitude

# Define the bounding box of the data
bbox <- bbox(data)

# Create a dense grid of points for prediction
grid <- expand.grid(
  Longitude = seq(bbox["Longitude", "min"], bbox["Longitude", "max"], length.out = 100),
  Latitude = seq(bbox["Latitude", "min"], bbox["Latitude", "max"], length.out = 100)
)

# Convert the grid to a spatial object
coordinates(grid) <- ~Longitude + Latitude
gridded(grid) <- TRUE

# Compute the empirical variogram
vario <- variogram(PM10 ~ 1, data)

# Fit a variogram model (e.g., exponential model)
vario_fit <- fit.variogram(vario, model = vgm(psill = 9000, model = "Exp", range = 15000, nugget = 1000))
if (any(vario_fit$range < 0)) stop("Fitted variogram has negative range — try different initial values")

# Plot the variogram and fitted model
plot(vario, vario_fit, main = "Variogram and Fitted Model")

# Perform Ordinary Kriging
ordinary_kriging <- krige(PM10 ~ 1, data, grid, model = vario_fit)

# Perform Universal Kriging (with a trend)
universal_kriging <- krige(PM10 ~ Longitude + Latitude, data, grid, model = vario_fit)

# Perform Simple Kriging (assuming a known mean)
simple_kriging <- krige(PM10 ~ 1, data, grid, model = vario_fit, beta = mean(data$PM10))

ordinary_cv <- krige.cv(PM10 ~ 1, data, model = vario_fit)
universal_cv <- krige.cv(PM10 ~ Longitude + Latitude, data, model = vario_fit)
simple_cv <- krige.cv(PM10 ~ 1, data, model = vario_fit, beta = mean(data$PM10))

set.seed(123)
ordinary_mse <- mean(ordinary_cv$residual^2)
universal_mse <- mean(universal_cv$residual^2)
simple_mse <- mean(simple_cv$residual^2)

ordinary_mse
universal_mse
simple_mse
```



```{r}
# Load necessary libraries
library(sp)
library(gstat)
library(ggplot2)

# Read the dataset
data <- read.csv("final_data.csv")

# Rename columns with special characters for reliable formula use
names(data)[names(data) == "Wind.Speed..U."] <- "WindU"
names(data)[names(data) == "Wind.Speed..V."] <- "WindV"
names(data)[names(data) == "Relative.Humidity"] <- "RelHum"

# Convert the data to a spatial object
coordinates(data) <- ~Longitude + Latitude

# Define the bounding box of the data
bbox <- bbox(data)

# Create a dense grid of points for prediction
grid <- expand.grid(
  Longitude = seq(bbox["Longitude", "min"], bbox["Longitude", "max"], length.out = 100),
  Latitude = seq(bbox["Latitude", "min"], bbox["Latitude", "max"], length.out = 100)
)

# Add predictor variables to the grid
# For simplicity, assume constant values for predictors across the grid
# (You can replace these with more sophisticated interpolation if needed)
grid$Temp <- mean(data$Temp)      # Use mean temperature
grid$RelHum <- mean(data$RelHum)  # Use mean relative humidity
grid$WindU <- mean(data$WindU)    # Use mean U-component wind speed
grid$WindV <- mean(data$WindV)    # Use mean V-component wind speed

# Convert the grid to a spatial object
coordinates(grid) <- ~Longitude + Latitude
gridded(grid) <- TRUE

# Compute the empirical variogram for PM10 with additional predictors
vario <- variogram(PM10 ~ Temp + RelHum + WindU + WindV, data)

# Fit a variogram model (e.g., exponential model)
# psill lowered to ~5000 as residual variogram (after trend removal) has smaller sill
vario_fit <- fit.variogram(vario, model = vgm(psill = 5000, model = "Exp", range = 15000, nugget = 1000))
if (any(vario_fit$range < 0)) stop("Fitted variogram has negative range — try different initial values")

# Perform Universal Kriging with additional predictors
universal_kriging_additional_predictors <- krige(
  PM10 ~ Temp + RelHum + WindU + WindV,  # Formula with predictors
  locations = data,  # Original data
  newdata = grid,    # Grid for predictions (with predictors)
  model = vario_fit  # Fitted variogram model
)

# Perform cross-validation for Universal Kriging with additional predictors
universal_cv_extra_predictors <- krige.cv(
  PM10 ~ Temp + RelHum + WindU + WindV,  # Formula with predictors
  data,  # Original data
  model = vario_fit  # Fitted variogram model
)

# Extract performance metrics
universal_extra_predictors_cv <- mean(universal_cv_extra_predictors$residual^2)

universal_extra_predictors_cv
```


```{r}
# Load necessary libraries
library(sp)
library(gstat)
library(ggplot2)
library(dplyr)

# Read the dataset
data <- read.csv("final_data.csv")

# Convert the data to a spatial object
coordinates(data) <- ~Longitude + Latitude

# Compute the empirical semivariogram
sample_variogram <- variogram(PM10 ~ 1, data)

# Fit spherical models with different fitting methods
variogram_ols <- fit.variogram(sample_variogram, vgm(model = "Sph"), fit.method = 0)  # OLS
variogram_npairs <- fit.variogram(sample_variogram, vgm(model = "Sph"), fit.method = 1)  # WLS (npairs)

# Fit exponential models with different fitting methods
exp_variogram_ols <- fit.variogram(sample_variogram, vgm(model = "Exp"), fit.method = 0)  # OLS
exp_variogram_npairs <- fit.variogram(sample_variogram, vgm(model = "Exp"), fit.method = 1)  # WLS (npairs)

# Perform cross-validation for each model
# Spherical models
cv_ols <- krige.cv(PM10 ~ 1, data, model = variogram_ols)
cv_npairs <- krige.cv(PM10 ~ 1, data, model = variogram_npairs)

# Exponential models
cv_exp_ols <- krige.cv(PM10 ~ 1, data, model = exp_variogram_ols)
cv_exp_npairs <- krige.cv(PM10 ~ 1, data, model = exp_variogram_npairs)

# Extract performance metrics for each model
metrics <- data.frame(
  Model = c("Spherical (OLS)", "Spherical (WLS npairs)",
            "Exponential (OLS)", "Exponential (WLS npairs)"),
  ME = c(mean(cv_ols$residual), mean(cv_npairs$residual),
        mean(cv_exp_ols$residual), mean(cv_exp_npairs$residual)),
  MSE = c(mean(cv_ols$residual^2), mean(cv_npairs$residual^2),
         mean(cv_exp_ols$residual^2), mean(cv_exp_npairs$residual^2)),
  RMSE = c(sqrt(mean(cv_ols$residual^2)), sqrt(mean(cv_npairs$residual^2)),
           sqrt(mean(cv_exp_ols$residual^2)), sqrt(mean(cv_exp_npairs$residual^2)))
)

# Print the performance metrics
print(metrics)
```

```{r}
# Load necessary libraries
library(sp)
library(gstat)
library(ggplot2)

# Read the dataset
data <- read.csv("final_data.csv")

# Rename columns with special characters for reliable formula use
names(data)[names(data) == "Wind.Speed..U."] <- "WindU"
names(data)[names(data) == "Wind.Speed..V."] <- "WindV"
names(data)[names(data) == "Relative.Humidity"] <- "RelHum"

# Convert the data to a spatial object
coordinates(data) <- ~Longitude + Latitude

# Define the bounding box of the data
bbox <- bbox(data)

# Create a dense grid of points for prediction
grid <- expand.grid(
  Longitude = seq(bbox["Longitude", "min"], bbox["Longitude", "max"], length.out = 100),
  Latitude = seq(bbox["Latitude", "min"], bbox["Latitude", "max"], length.out = 100)
)

# Add predictor variables to the grid
# For simplicity, assume constant values for predictors across the grid
# (You can replace these with more sophisticated interpolation if needed)
grid$Temp <- mean(data$Temp)      # Use mean temperature
grid$RelHum <- mean(data$RelHum)  # Use mean relative humidity
grid$WindU <- mean(data$WindU)    # Use mean U-component wind speed
grid$WindV <- mean(data$WindV)    # Use mean V-component wind speed

# Convert the grid to a spatial object
coordinates(grid) <- ~Longitude + Latitude
gridded(grid) <- TRUE

# Compute the empirical variogram for PM10 with additional predictors
vario <- variogram(PM10 ~ Temp + RelHum + WindU + WindV, data)

# Fit the Exponential (WLS npairs) model to the empirical variogram
vario_fit <- fit.variogram(vario, model = vgm(psill = 5000, model = "Exp", range = 15000, nugget = 1000), fit.method = 1)  # WLS (npairs)
if (any(vario_fit$range < 0)) stop("Fitted variogram has negative range — try different initial values")

# Print the fitted variogram model
print(vario_fit)

# Perform Universal Kriging with additional predictors
universal_kriging_additional_predictors <- krige(
  PM10 ~ Temp + RelHum + WindU + WindV,  # Formula with predictors
  locations = data,  # Original data
  newdata = grid,    # Grid for predictions (with predictors)
  model = vario_fit  # Fitted variogram model
)

# Convert the results to a data frame for plotting
universal_df <- as.data.frame(universal_kriging_additional_predictors)

# Plot the results
# Clamp colour scale to 5th–95th percentile of predictions to reveal spatial variation
pred_range <- quantile(universal_df$var1.pred, c(0.05, 0.95), na.rm = TRUE)

ggplot(universal_df, aes(x = Longitude, y = Latitude, fill = var1.pred)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", name = "PM10 Concentration",
                       limits = pred_range, oob = scales::squish) +
  labs(title = "Universal Kriging Predictions for PM10 (with Additional Predictors)",
       x = "Longitude",
       y = "Latitude") +
  theme_minimal()

# Perform cross-validation for Universal Kriging with additional predictors
universal_cv_extra_predictors <- krige.cv(
  PM10 ~ Temp + RelHum + WindU + WindV,  # Formula with predictors
  data,  # Original data
  model = vario_fit  # Fitted variogram model
)

# Extract performance metrics
universal_extra_predictors_cv <- mean(universal_cv_extra_predictors$residual^2)

# Print the cross-validation MSE
print(paste("Cross-Validation MSE:", universal_extra_predictors_cv))
```


