---
title: "Optimal Pre-Workout Drink and Dosage: Two-Way ANOVA Analysis"
subtitle: "STATS 101B Final Project"
author: "Jeremy Shiu, Derek Nakagawa, Luke Shen, Kai Zheng"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
---

```{r setup, include=FALSE}
# Global chunk options:
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

Code for my STATS 101B project on whether a **pre-workout drink** (water, sugar water, or a sugar-free energy drink) and its **dosage** (250, 500, or 750 mL) affect exercise performance.

The experiment is a **balanced two-factor factorial design** (3 Drinks x
3 Dosages = 9 treatment groups, 28 subjects each, 252 total) analyzed with a
**two-way ANOVA with interaction**. The response is the *percent change* in the
number of bicep curls a subject can perform in 30 seconds, comparing a
no-drink baseline to a post-treatment measurement:

$$\text{Difference} = \frac{X_2 - X_1}{X_1}$$

where \(X_1\) = baseline reps and \(X_2\) = reps after the assigned treatment.
Using percent change (rather than raw reps) controls for individual variation
in baseline strength.

# Packages

```{r packages}
# ggplot2 : all boxplots / interaction plots
# dplyr   : group_by() + summarise() for the interaction-plot summary table
# pwr     : a-priori sample size / power calculation
library(ggplot2)
library(dplyr)
library(pwr)
```

# Load and Prepare the Data

```{r load-data}
raw <- read.csv("data for stats 101b project - Sheet1.csv",
                stringsAsFactors = FALSE)

# turn "Dosage(mL)" into "Dosage.mL." -- rename for convenience.
names(raw)[names(raw) == "Dosage.mL."] <- "Dosage"

dat <- data.frame(
  Baseline   = raw$Baseline,       # X1: baseline reps (no drink)
  Drink      = raw$Drink,          # Factor 1: water / sugar water / sugar free energy drink
  Dosage     = raw$Dosage,         # Factor 2: 250 / 500 / 750 mL
  Reps       = raw$Repetitions,    # X2: reps after treatment
  Difference = raw$Difference      # Response: (X2 - X1) / X1
)

# --- Convert the two treatment columns to FACTORS -------------------------
# Dosage is numeric in the CSV; we treat it as a categorical factor because
# the design has three discrete dose levels, not a continuous predictor.
dat$Drink  <- factor(dat$Drink,
                     levels = c("water", "sugar water", "sugar free energy drink"))
dat$Dosage <- factor(dat$Dosage, levels = c(250, 500, 750))

# Quick sanity checks: structure and the balance of the design.
str(dat)
table(dat$Drink, dat$Dosage)   # should be 28 in every cell (balanced design)
summary(dat$Difference)
```

# Sample Size Determination

```{r sample-size}
# A-priori power analysis done BEFORE collecting data.
#   - Desired power : 0.80
#   - Significance  : alpha = 0.05
#   - Effect size   : Cohen's medium effect, f = 0.25  (f^2 = 0.0625)
#
# For a factorial ANOVA the relevant test is an F-test whose NUMERATOR df (u)
# equals the df of the effect being powered. pwr.f2.test returns the
# DENOMINATOR df (v); total N = u + v + 1.
#
# NOTE (documented in the report): we originally used u = 8 (the df you would
# get by treating all 9 groups as one factor, 9 - 1 = 8), which gave ~249 and
# was rounded up to 252 for a balanced 28-per-cell design. The interaction term
# we actually care about has u = 4, so u = 8 slightly OVER-samples -- which is
# harmless (extra power), just not minimal.

# Reproduce the number reported in the paper (u = 8):
ss_reported <- pwr.f2.test(u = 8, f2 = 0.25^2, sig.level = 0.05, power = 0.80)
ss_reported
ceiling(ss_reported$v) + 8 + 1   # total N implied (~249 -> rounded to 252)

# The "correct" calculation for the interaction df (u = 4), for reference:
pwr.f2.test(u = 4, f2 = 0.25^2, sig.level = 0.05, power = 0.80)
```

# Exploratory Plots

Before any formal testing we look at the data to form expectations.

## Boxplots

```{r boxplot-drink}
# Response distribution by DRINK type.
# A reference band of +/- 0.10 is drawn because a 10% change would be
# meaningful to an athlete even though the raw spread looks small.
ggplot(dat, aes(x = Drink, y = Difference, fill = Drink)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(title = "Percent Change in Reps by Drink",
       x = "Drink", y = "Percent Difference") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r boxplot-dosage}
# Response distribution by DOSAGE.
ggplot(dat, aes(x = Dosage, y = Difference, fill = Dosage)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(title = "Percent Change in Reps by Dosage",
       x = "Dosage (mL)", y = "Percent Difference") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r boxplot-both}
# Combined view: dosage on the x-axis, drink shown by fill.
# This is the clearest single picture of all 9 treatment groups.
ggplot(dat, aes(x = Dosage, y = Difference, fill = Drink)) +
  geom_boxplot(alpha = 0.8, position = position_dodge(0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  labs(title = "Percent Change by Drink and Dosage",
       x = "Dosage (mL)", y = "Percent Difference", fill = "Drink") +
  theme_minimal()
```

The between-group differences in means are small relative to the within-group
spread, so we do not expect large main effects.

## Interaction Plot

```{r interaction-plot}
# Build the cell-mean summary table with dplyr (group_by + summarise),
# exactly as described in the report, then plot the means with ggplot.
cell_means <- dat %>%
  group_by(Drink, Dosage) %>%
  summarise(mean_diff = mean(Difference), .groups = "drop")

ggplot(cell_means, aes(x = Dosage, y = mean_diff,
                       color = Drink, group = Drink)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  labs(title = "Interaction Plot: Mean Percent Difference",
       x = "Dosage (mL)", y = "Mean Percent Difference", color = "Drink") +
  theme_minimal()

# Base-R equivalent (produces the classic interaction.plot layout):
with(dat, interaction.plot(x.factor = Dosage, trace.factor = Drink,
                           response = Difference, fun = mean,
                           xlab = "Dosage (mL)", ylab = "Mean Percent Difference",
                           trace.label = "Drink", col = 1:3, lwd = 2))
```

The non-parallel lines (especially the sugar-free energy drink rising then
falling as dose increases) suggest a possible interaction -- so we expect the
interaction term to be the closest to significant in the ANOVA.

# Two-Way ANOVA (Full Model)

```{r full-anova}
# Full factorial model: two main effects + their interaction.
# The formula "Dosage * Drink" expands to  Dosage + Drink + Dosage:Drink.
# We list Dosage first so the sequential (Type I) ANOVA table matches the
# ordering in the report (Dosage, then Drink, then the interaction).
model_full <- aov(Difference ~ Dosage * Drink, data = dat)
summary(model_full)
```

This reproduces the reported ANOVA table:

| Source        | Df  | Sum Sq | Mean Sq  | F     | p       |
|---------------|-----|--------|----------|-------|---------|
| Dosage        | 2   | 0.0047 | 0.002347 | 0.294 | 0.7458  |
| Drink         | 2   | 0.0207 | 0.010370 | 1.298 | 0.2750  |
| Dosage:Drink  | 4   | 0.0631 | 0.015786 | 1.975 | 0.0989  |
| Residuals     | 243 | 1.9419 | 0.007991 |       |         |

Neither main effect is significant (p >> 0.05). The interaction is the closest
to significant (p = 0.0989) but still above alpha = 0.05.

# Reduced Models

```{r reduced-models}
# The interaction was closest to significant, so we probe a few reduced models
# to see whether any isolates a significant effect. In R, "A:B" is the
# interaction term only; "A + A:B" keeps one main effect plus the interaction.

# (a) Dosage + interaction
model_dose_int <- aov(Difference ~ Dosage + Dosage:Drink, data = dat)
summary(model_dose_int)

# (b) Drink + interaction
model_drink_int <- aov(Difference ~ Drink + Dosage:Drink, data = dat)
summary(model_drink_int)

# (c) Interaction only
model_int_only <- aov(Difference ~ Dosage:Drink, data = dat)
summary(model_int_only)
```

None of the reduced models produces a significant p-value, confirming that the
full model is not hiding a significant effect. The full factorial model remains
our final model.

# Model Diagnostics

```{r diagnostics, fig.width=8, fig.height=7}
# The four standard diagnostic plots for the full model:
#   1. Residuals vs Fitted  -> checks constant variance / no pattern
#   2. Normal Q-Q           -> checks normality of residuals
#   3. Scale-Location       -> checks homoscedasticity (spread of residuals)
#   4. Residuals vs Factor Levels / Constant Leverage -> influential points
par(mfrow = c(2, 2))   # 2x2 grid
plot(model_full)
par(mfrow = c(1, 1))   # reset plotting layout
```

The residuals scatter randomly around zero with roughly constant spread, and
the Q-Q plot is close to the reference line (one mild outlier). Both the
constant-variance and normality assumptions hold, so the model is valid.

# Tukey HSD (Post-Hoc)

```{r tukey}
# Even though no factor was significant, we report Tukey's Honest Significant
# Difference for completeness (pairwise comparisons within each main effect).
TukeyHSD(model_full, which = c("Dosage", "Drink"))
```

As expected from the non-significant F-tests, every adjusted p-value is large,
so no pair of dosages or drinks differs significantly.

# Conclusions

Across all tests, **no factor reached significance**:

- **Dosage** (p = 0.7458) and **Drink** (p = 0.2750) main effects are far from
  significant.
- The **Drink x Dosage interaction** (p = 0.0989) is the closest to
  significant but still above alpha = 0.05, and reduced models do not recover a
  significant effect.
- The sugar-free energy drink shows the strongest *positive* mean effect on
  bicep curls (consistent with the caffeine literature), but the effect is too
  small to be statistically significant.
- Diagnostic plots confirm the model assumptions hold, so these conclusions are
  valid.

We therefore conclude that, in this experiment, the type of drink, its dosage,
and their interaction had **no statistically significant effect** on bicep-curl
performance.

